The Multi-Stencil Language

### 2016-01-22 ARTICLES/IJPP/article/formalism.tex
article IJPP add reduction Helene 22/01/2016

\begin{mydef}
A computation kernel $k$ of a numerical simulation is defined as $k(R,(w,D),comp)$, where 
\begin{itemize}
\item $(w,D) \in \Delta \times \mathcal{D}$ is the single data modified by the computation kernel onto a given computation domain. Thus, $w \in \Delta$ and $D \in \mathcal{D}$ is the computation domain on which $w$ is computed, $D \subset E_w=entity(w)$.
\item $R$ is the set of tuples $(r,n)$, where $r \in \Delta$ and $n \in \mathcal{N}$ is a neighborhood function such that $n : E_w \rightarrow entity(r)^m$. The neighborhood indicates which entities of $r$ will be read in order to compute a single entity of $w$. 
\item $comp$ is the numerical computation which returns a value from a set of input values, $comp: V^n \rightarrow V$.
\end{itemize}
% \item A numerical expression $\text{exp}_{w}: entity(w) \times R^n \rightarrow V_{w}$ is a function which computes the values of the written quantity $w \in \Delta$ for a given entity $e \in entity(w)=E_w$, using a set of input data $R$.
% \item A computation kernel $k$ of a numerical simulation is defined as $c(R,w,\text{exp},D)$, where $R$ is the set of data read, $w \in \Delta$ is the unique data written, $\text{exp}$ is a numerical expression, and $D \in \mathcal{D}$ is the computation domain.
% \end{itemize}
%$R$ is the set of data read, $w \in \Delta$ is the unique data written, $\text{exp}$ is a numerical expression, and $D \in \mathcal{D}$ is the computation domain.
\end{mydef}

@@

\noindent \textbf{Definitions}
%\begin{mydef}
\begin{itemize}
\item We denote by $identity$ the identity function $x=x$.
\item A kernel computation $k(R,(w,D),comp)$ is a \emph{stencil kernel} $\iff \exists (r,n) \in R$ such that $n \neq identity$.
\item A kernel computation $k(R,(w,D),comp)$ is a \emph{local kernel} $\iff \forall (r,n) \in R$, $n = identity$.
\item A kernel computation $k(R,(w,D),comp)$ is a \emph{reduction kernel} if $w$ is a scalar, in other words $\iff dim(mesh(entity(w)))=0$.
\end{itemize}

@@

\medskip
A reduction computation is a computation which takes data applied onto meshes and returns a data with a dimension reduced to zero, in other words a scalar. A reduction is typically used to compute the convergence criteria of the time loop of the simulation. Occasionally reductions can also be performed during a time iteration. %This last usage is typically done to have a conditionnal branch to choose one computation or another, or to compute a dynamic time step for example.

\noindent \textbf{Property}
Considering a reduction kernel $k(R,(w,D),comp)$, $comp$ must be a binary and associative operation on the type $V$, $comp: V \times V \rightarrow V$.

@@

A stencil program has been formally defined in this section. This formalism is used in the next Section to define two mid-grain parallelization techniques of a multi-stencil program.

>>>>

\begin{mydef}
A computation kernel $k$ of a numerical simulation is defined as $k(S,R,(w,D),comp)$, where 
\begin{itemize}
\item $S \in \mathcal{S}$ is the set of scalar to read, 
\item $(w,D) \in \Delta \times \mathcal{D}$ is the single data modified by the computation kernel onto a given computation domain. Thus, $w \in \Delta$ and $D \in \mathcal{D}$ is the computation domain on which $w$ is computed, $D \subset E_w=entity(w)$.
\item $R$ is the set of tuples $(r,n)$, where $r \in \Delta$ and $n \in \mathcal{N}$ is a neighborhood function such that $n : E_w \rightarrow entity(r)^m$. The neighborhood indicates which entities of $r$ will be read in order to compute a single entity of $w$. 
\item $comp$ is the numerical computation which returns a value from a set of input values, $comp: V^n \rightarrow V$.
\end{itemize}
% \item A numerical expression $\text{exp}_{w}: entity(w) \times R^n \rightarrow V_{w}$ is a function which computes the values of the written quantity $w \in \Delta$ for a given entity $e \in entity(w)=E_w$, using a set of input data $R$.
% \item A computation kernel $k$ of a numerical simulation is defined as $c(R,w,\text{exp},D)$, where $R$ is the set of data read, $w \in \Delta$ is the unique data written, $\text{exp}$ is a numerical expression, and $D \in \mathcal{D}$ is the computation domain.
% \end{itemize}
%$R$ is the set of data read, $w \in \Delta$ is the unique data written, $\text{exp}$ is a numerical expression, and $D \in \mathcal{D}$ is the computation domain.
\end{mydef}

@@

\noindent \textbf{Definitions}
%\begin{mydef}
\begin{itemize}
\item We denote by $identity$ the identity function $x=x$.
\item A kernel computation $k(S,R,(w,D),comp)$ is a \emph{stencil kernel} $\iff \exists (r,n) \in R$ such that $n \neq identity$.
\item A kernel computation $k(S,R,(w,D),comp)$ is a \emph{local kernel} $\iff \forall (r,n) \in R$, $n = identity$.
\item A kernel computation $k(S,R,(w,D),comp)$ is a \emph{reduction kernel} $\iff w$ is a scalar, $R=\{(r,n)\}$, and if $n=entity(r)$.
\end{itemize}

@@

\noindent \textbf{Property}
In a reduction kernel $k(S,R,(w,D),comp)$, $D=entity(w)$ as a single entity exists for a scalar.

\medskip
A reduction computation is a computation which takes a single data applied onto a mesh and returns from all its entities a single data with a dimension reduced to zero, in other words a scalar. A reduction is typically used to compute the convergence criteria of the time loop of the simulation. Occasionally reductions can also be performed during a time iteration. %This last usage is typically done to have a conditionnal branch to choose one computation or another, or to compute a dynamic time step for example.

\noindent \textbf{Property}
Considering a reduction kernel $k(S,R,(w,D),comp)$, $comp$ must be a binary and associative operation on the type $V$, $comp: V \times V \rightarrow V$.

@@

A stencil program has been formally defined in this section. This formalism is used in the next Section to define two parallelization techniques of a multi-stencil program.

### 2016-01-22 ARTICLES/IJPP/article/parallelism.tex
article IJPP add reduction Helene 22/01/2016

More formally, the data parallelization of a multi-stencil program $\mathcal{MSP}(T,\mathcal{M},\mathcal{E},\mathcal{D},\Delta, \mathcal{S},\Gamma)$ consists in, first, a partitioning of the mesh $\mathcal{M}$ in $p$ balanced sub-meshes (for $p$ resources) $\mathcal{M}=\{\mathcal{M}_0,\dots,\mathcal{M}_{p-1}\}$. This step can be performed by an external graph partitionner~\cite{} and is not adressed by this paper. As a data is mapped onto the mesh, the set of data $\Delta$ is partitionned the same way than the mesh in $\Delta=\{\Delta_0,\dots,\Delta_{p-1}\}$. The second step of the parallelization is to identify in $\Gamma$ the needed synchronizations between resources to update data, and thus to build a new ordered list of computations $\Gamma_{sync}$.

\begin{mydef}
\textit{
For $n$ the number of computations in $\Gamma$, and for $i,j$ such that $i<j<n$, a \textit{synchronisation} is needed between $k_i$ and $k_j$, denoted $k_i \pprec k_j$, if $\exists (r,n) \in R_j$ with $n\neq identity$ ($k_j$ is a stencil computation), and such that $\{w_i\} \subset R_j$. Moreover, the data to update is $\{w_i\} \cap R_j = \{w_i$\}.}
\end{mydef}

Actually, a synchronization can only be needed by the data read by a stencil computation (not local), and only if this data has been modified before, which means that it has been written before. This synchronization is needed because a neighborhood function $n \in \mathcal{N}$ of a stencil computation involves values computed on different resources.

\begin{mydef}
\textit{A synchronization between two computations $k_i \pprec k_j$ is defined as a specific computation $sync(R,w,exp,D)$ where $R=\{w_i\}$, $w=w_i$, $exp=identity$ and $D=\emptyset$.}
\end{mydef}

\begin{mydef}
If a synchronization is needed before $k_j$ (on the data $w_i$), $k_j$ is replaced in $\Gamma_{sync}$ by the list
\begin{equation*}
\Gamma_{sync} = [sync(\{w_i\},w_i,identity,\emptyset), k_j]
\end{equation*}
\end{mydef}

@@

\begin{mydef}
From the ordered list of computation $\Gamma$, a new ordered list $\Gamma_{sync}$ is obtained from the call $\Gamma_{sync} = F_{sync}(\Gamma,0)$, where $F_{sync}$ is the recursive function defined as
\begin{equation*}
F_{sync}(\Gamma,j) = 
\begin{cases} 	[\Gamma[j]] \cdot F_{sync}(\Gamma,j+1) & \mbox{if }\forall i<j \mbox{, } k_i \not\pprec k_j\\
				[sync(\{w_i\},w_i,identity,\emptyset), k_j] \cdot F_{sync}(\Gamma,j+1) & \forall i<j \text{, and }k_i \pprec k_j\}\\
				[] & \mbox{if }j=|\Gamma|.
\end{cases}
\end{equation*}
\end{mydef}

@@

 The final step of this parallelization is to run $\Gamma_{sync}$ on each resource. Thus, for each resource $0 \leq k \leq p-1$ a multi-stencil program, defined by
\begin{equation}
\mathcal{MSP}_k(T,\mathcal{M}_k,\mathcal{E},\mathcal{D},\Delta_k, \mathcal{S},\Gamma_{sync}),
\end{equation}
is performed.

>>>>

More formally, the data parallelization of a multi-stencil program $\mathcal{MSP}(T,\mathcal{M},\mathcal{E},\mathcal{D},\Delta, \mathcal{S},\Gamma)$ consists in, first, a partitioning of the mesh $\mathcal{M}$ in $p$ balanced sub-meshes (for $p$ resources) $\mathcal{M}=\{\mathcal{M}_0,\dots,\mathcal{M}_{p-1}\}$. This step can be performed by an external graph partitionner~\cite{} and is not adressed by this paper. As entities and as data are mapped onto the mesh, the set of entities and the set of data $\Delta$ are partitionned the same way than the mesh: $\mathcal{E}=\{\mathcal{E},\dots,\mathcal{E}{p-1}\}$, $\Delta=\{\Delta_0,\dots,\Delta_{p-1}\}$. The second step of the parallelization is to identify in $\Gamma$ the needed synchronizations between resources to update data, and thus to build a new ordered list of computations $\Gamma_{sync}$.

\begin{mydef}
For $n$ the number of computations in $\Gamma$, and for $i,j$ such that $i<j<n$, a \textit{synchronisation} is needed between $k_i$ and $k_j$, denoted $k_i \pprec k_j$, if $\exists (r_j,n_j) \in R_j$ such that $w_i=r_j$ and $n_j\neq identity$ ($k_j$ is a stencil computation). Moreover, the data to update is $\{w_i\}$.
\end{mydef}

Actually, a synchronization is needed by the data read by a stencil computation (not local), and only if this data has been modified before, which means that it has been written before. This synchronization is needed because a neighborhood function $n \in \mathcal{N}$ of a stencil computation involves values computed on different resources.

\begin{mydef}
A synchronization between two computations $k_i \pprec k_j$ is defined as a specific computation 
\begin{equation*}
sync(S,R,w,comp)
\end{equation*}
\end{mydef}
where $S=\emptyset$, $R=\{(w_i,n_j)\}$, $w=(w_i,n_j)$ and $comp=identity$. In other words, $w_i$ has to be synchronized for the neighborhood $n_j$.

\begin{mydef}
If $k_i \pprec k_j$, $k_j$ is replaced by the list
\begin{equation*}
[sync(\emptyset,\{(w_i,n_j)\},(w_i,n_j),identity), k_j]
\end{equation*}
\end{mydef}

When a data parallelism is applied, the other type of computation which is responsible for additional synchronizations is the reduction. Actually, the reduction is applied locally on each subset of entities, on each resource, which results in $p$ (number of resources) scalar value, one on each resource. Thus, to perform the final reduction a synchronization is needed to get the $p$ scalars onto a single resource, to finally get the reduction scalar.

\begin{mydef}
For a reduction kernel $k(S,R,(w,D),comp)$, if we denote by $w^i$, $0 \leq i<p$, the local $w$ computed on the resource $i$, a reduction synchronization is defined as the specific computation 
\begin{equation*}
sync(S,R_s,w,comp)
\end{equation*}
\end{mydef}
where, $S=\emptyset$, $R=\{(w^0,entity(w^0)) \dots (w^{p-1},entity(w^{p-1}))\}$ and $w$ and $comp$ are the same than the actual reduction.

One can notice that both types of synchronizations are performed by all resources.

@@

\begin{mydef}
From the ordered list of computation $\Gamma$, a new ordered list $\Gamma_{sync}$ is obtained from the call $\Gamma_{sync} = F_{sync}(\Gamma,0)$, where $F_{sync}$ is the recursive function defined as
\begin{equation*}
F_{sync}(\Gamma,j) = 
\begin{cases} 	[] \mbox{ if }j=|\Gamma|\\
				[sync(\emptyset,\{w_i\},w_i,identity,\emptyset), k_j] \cdot F_{sync}(\Gamma,j+1)\\
				\text{ } \qquad \forall i<j \text{, and }k_i \pprec k_j\\
				[k_j, sync(\emptyset,,w_j,comp_j)] \cdot F_{sync}(\Gamma,j+1)\\ 
				\text{ } \qquad \mbox{if } w_j \in \mathcal{S} \text{, and } R_j=\{(r_j,entity(r_j))\}\\
				[\Gamma[j]] \cdot F_{sync}(\Gamma,j+1) \text{ else}.
\end{cases}
\end{equation*}
\end{mydef}

@@

 The final step of this parallelization is to run $\Gamma_{sync}$ on each resource. Thus, for each resource $0 \leq k \leq p-1$ a multi-stencil program, defined by
\begin{equation}
\mathcal{MSP}_k(T,\mathcal{M}_k,\mathcal{E},\mathcal{D},\Delta_k,\mathcal{S},\Gamma_{sync}),
\end{equation}
is performed.
